---
title: "Advent of Code 2019"
subtitle: "Day 2"
author: "Romain Lesur"
date: "02/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

My first idea is to use a reference class object. I feel that a closure could suffice.

## First step: process the input.

```{r}
#' Incode computer
#'
#' @param program A string with comma separated integers (an Intcodes program)
#'
#' @return An object
computer <- function(program) {
  program <- strsplit(program, ",")[[1]] # use [[1]] because we have a single program
  program <- as.integer(program)
  
  list(
    current_program = function() { program }
  )
}
```

OK. Let's test that.

```{r}
program <- "1,9,10,3,2,3,11,0,99,30,40,50"

process <- computer(program)

process$current_program()
```

It seems to be OK.

Oh! I have a fun idea to process the input! The program is very similar to a vector input in R. If I add "c(" before the program and ")" after, this is how I write a vector. So, I want to try with `parse()`.

```{r}
computer <- function(program) {
  program <- eval(parse(text = paste0("c(", program, ")")))
  program <- as.integer(program)
  
  list(
    current_program = function() { program }
  )
}
```

Now, test that.

```{r}
program <- "1,9,10,3,2,3,11,0,99,30,40,50"

process <- computer(program)

process$current_program()
```

That's good. I like this trick. I want to keep it.

## Reading a stack

OK. Now, I have to deal with the positions and reading a stack of 4 integers. Let's keep in mind that vectors are indexed from 1 in R and not 0. This is a trap for this game.

```{r}
computer <- function(program) {
  program <- eval(parse(text = paste0("c(", program, ")")))
  program <- as.integer(program)
  
  pointer <- 0L # I want to respect the convention defined in the game
  
  read_position <- function(position) { 
    program[position + 1L] 
  }
  
  read_stack <- function(position) {
      stack_positions <- seq.int(position + 1L, position + 4L)
      program[stack_positions]
  }
  
  list(
    current_program = function() { program },
    get_pointer_position = function() { pointer },
    get_pointer_value = function() { read_position(pointer) },
    get_stack_from_pointer = function() { read_stack(pointer) },
    move_pointer = function() { pointer <<- pointer + 4L }
  )
}
```

Now, I want to test this new methods

```{r}
program <- "1,9,10,3,2,3,11,0,99,30,40,50"

process <- computer(program)

process$current_program()

test_identical <- function(...) stopifnot(identical(...))

test_identical(process$get_pointer_position(), 0L)

test_identical(process$get_pointer_value(), 1L)

test_identical(
  process$get_stack_from_pointer(), 
  c(1L, 9L, 10L, 3L)
)

process$move_pointer()

test_identical(process$get_pointer_position(), 4L)
test_identical(process$get_pointer_value(), 2L)
test_identical(
  process$get_stack_from_pointer(), 
  c(2L, 3L, 11L, 0L)
)
```

All tests passed: there's no error. 

## Perform the operation

Now, let's add some methods to perform the operation and override values.

```{r}
computer <- function(program) {
  program <- eval(parse(text = paste0("c(", program, ")")))
  program <- as.integer(program)
  
  pointer <- 0L # I want to respect the convention defined in the game
  
  read_position <- function(position) { 
    program[position + 1L] 
  }
  
  read_stack <- function(position) {
      stack_positions <- seq.int(position + 1L, position + 4L)
      program[stack_positions]
  }
  
  # I move get_pointer_value() here for further reuse
  get_pointer_value <- function() { 
    read_position(pointer) 
  }

  # I create a new function to get the operation
  get_operation <- function() {
    opcode <- get_pointer_value()
    switch(opcode, sum, prod)
  }
  
  # I move get_stack_from_pointer here for further reuse
  get_stack_from_pointer <- function() { 
    read_stack(pointer) 
  }
  
  # I create a new function to get the operands
  get_operands <- function() {
    current_stack <- get_stack_from_pointer()
    positions <- current_stack[2:3]
    read_position(positions)
  }
  
  # I create a new function to compute the stack
  compute_value_from_current_stack <- function() {
    operation <- get_operation()
    operands <- get_operands()
    operation(operands)
  }
  
  # I create a new function to overwrite a value
  overwrite <- function(position, value) {
    new_program <- program # init
    new_program[position + 1L] <- value
    program <<- new_program # note: <<- is used. That was the motivation for a closure.
  }

  # a new function to get the output position
  get_output_position <- function() {
    current_stack <- get_stack_from_pointer()
    current_stack[4]
  }
  
  # a new function to compute and overwrite the target position
  overwrite_output_position <- function() {
    overwrite(get_output_position(), compute_value_from_current_stack())
  }
  
  list(
    current_program = function() { program },
    get_pointer_position = function() { pointer },
    get_pointer_value = get_pointer_value, # moved
    get_stack_from_pointer = get_stack_from_pointer, # moved
    move_pointer = function() { pointer <<- pointer + 4L },
    get_operands = get_operands, # new method
    compute_value_from_current_stack = compute_value_from_current_stack, # new method
    get_output_position = get_output_position, # new method
    overwrite_output_position = overwrite_output_position # new method
  )
}
```

Now, some tests.  
First, let's keep the previous tests.

```{r}
program <- "1,9,10,3,2,3,11,0,99,30,40,50"

process <- computer(program)

process$current_program()

test_identical <- function(...) stopifnot(identical(...))

test_identical(process$get_pointer_position(), 0L)

test_identical(process$get_pointer_value(), 1L)

test_identical(
  process$get_stack_from_pointer(), 
  c(1L, 9L, 10L, 3L)
)

process$move_pointer()

test_identical(process$get_pointer_position(), 4L)
test_identical(process$get_pointer_value(), 2L)
test_identical(
  process$get_stack_from_pointer(), 
  c(2L, 3L, 11L, 0L)
)
```

Add more tests for the new methods.

```{r}
program <- "1,9,10,3,2,3,11,0,99,30,40,50"

# Warning: we moved the pointer in the previous test
# Let's take a new fresh process
process <- computer(program)

test_identical(process$get_output_position(), 3L)
test_identical(process$compute_value_from_current_stack(), 70L)
process$overwrite_output_position()
test_identical(
  process$current_program(),
  c(1L, 9L, 10L, 70L, 2L, 3L, 11L, 0L, 99L, 30L, 40L, 50L)
)
# Now, let's take the second operation:
process$move_pointer()
process$overwrite_output_position()
```

**Now, I have a problem. The following test fails:**

```{r, eval=FALSE}
test_identical(
  process$current_program(),
  c(3500L, 9L, 10L, 70L, 2L, 3L, 11L, 0L, 99L, 30L, 40L, 50L)
)
```

The problem comes from the `prod()` function which does not return an integer.

I have to modify `compute_value_from_current_stack()` to coerce to an integer.

```{r}
computer <- function(program) {
  program <- eval(parse(text = paste0("c(", program, ")")))
  program <- as.integer(program)
  
  pointer <- 0L # I want to respect the convention defined in the game
  
  read_position <- function(position) { 
    program[position + 1L] 
  }
  
  read_stack <- function(position) {
      stack_positions <- seq.int(position + 1L, position + 4L)
      program[stack_positions]
  }
  
  get_pointer_value <- function() { 
    read_position(pointer) 
  }

  get_operation <- function() {
    opcode <- get_pointer_value()
    switch(opcode, sum, prod)
  }
  
  get_stack_from_pointer <- function() { 
    read_stack(pointer) 
  }
  
  get_operands <- function() {
    current_stack <- get_stack_from_pointer()
    positions <- current_stack[2:3]
    read_position(positions)
  }
  
  compute_value_from_current_stack <- function() {
    operation <- get_operation()
    operands <- get_operands()
    as.integer(operation(operands)) # HERE IS THE PROBLEM
  }
  
  overwrite <- function(position, value) {
    new_program <- program # init
    new_program[position + 1L] <- value
    program <<- new_program # note: <<- is used. That was the motivation for a closure.
  }

  get_output_position <- function() {
    current_stack <- get_stack_from_pointer()
    current_stack[4]
  }
  
  overwrite_output_position <- function() {
    overwrite(get_output_position(), compute_value_from_current_stack())
  }
  
  list(
    current_program = function() { program },
    get_pointer_position = function() { pointer },
    get_pointer_value = get_pointer_value,
    get_stack_from_pointer = get_stack_from_pointer,
    move_pointer = function() { pointer <<- pointer + 4L },
    get_operands = get_operands,
    compute_value_from_current_stack = compute_value_from_current_stack,
    get_output_position = get_output_position,
    overwrite_output_position = overwrite_output_position
  )
}
```

Now, the tests for computing:

```{r}
program <- "1,9,10,3,2,3,11,0,99,30,40,50"

# Let's take a new fresh process
process <- computer(program)

test_identical(process$get_output_position(), 3L)
test_identical(process$compute_value_from_current_stack(), 70L)
process$overwrite_output_position()
test_identical(
  process$current_program(),
  c(1L, 9L, 10L, 70L, 2L, 3L, 11L, 0L, 99L, 30L, 40L, 50L)
)
# Now, let's take the second operation:
process$move_pointer()
process$overwrite_output_position()
test_identical(
  process$current_program(),
  c(3500L, 9L, 10L, 70L, 2L, 3L, 11L, 0L, 99L, 30L, 40L, 50L)
)
```

It works. Cool!

Now, I have to implement a loop and halt the program when `99L` is encountered.

## Halt and catch fire

I haven't implemented yet a function to check the _opcode_. This check will be performed at the initialization and after the pointer has been moved. I think I have to implement some current state values. 

I think that two current states would be useful:

- a logical indicating if the current stack waits for computation
- a logical indicating if the computer is running

```{r}
computer <- function(program) {
  # new state variables
  waiting_for_computation <- TRUE # init
  running <- TRUE # init
  
  program <- eval(parse(text = paste0("c(", program, ")")))
  program <- as.integer(program)
  
  pointer <- 0L # I want to respect the convention defined in the game
  
  read_position <- function(position) { 
    program[position + 1L] 
  }
  
  read_stack <- function(position) {
      stack_positions <- seq.int(position + 1L, position + 4L)
      program[stack_positions]
  }
  
  get_pointer_value <- function() { 
    read_position(pointer) 
  }

  # NEW method
  check_opcode <- function() {
    opcode <- get_pointer_value()
    if (!(opcode %in% c(1L, 2L, 99L))) {
      stop("Wrong Opcode. Read ", opcode, " at position ", pointer, 
           ". Opcode must be 1, 2 or 99.")
    }
    
    if (identical(opcode, 99L)) {
      running <<- FALSE
      waiting_for_computation <<- FALSE
    }
  }
  
  # NEW We must check the first opcode during the initialisation
  check_opcode()
  
  get_operation <- function() {
    opcode <- get_pointer_value()
    switch(opcode, sum, prod)
  }
  
  get_stack_from_pointer <- function() { 
    read_stack(pointer) 
  }
  
  get_operands <- function() {
    current_stack <- get_stack_from_pointer()
    positions <- current_stack[2:3]
    read_position(positions)
  }
  
  compute_value_from_current_stack <- function() {
    operation <- get_operation()
    operands <- get_operands()
    as.integer(operation(operands))
  }
  
  overwrite <- function(position, value) {
    new_program <- program # init
    new_program[position + 1L] <- value
    program <<- new_program # note: <<- is used. That was the motivation for a closure.
  }

  get_output_position <- function() {
    current_stack <- get_stack_from_pointer()
    current_stack[4]
  }
  
  overwrite_output_position <- function() {
    # NEW Control the state
    if (!isTRUE(running)) {
      stop("Cannot compute: program halted.")
    }
    if (!isTRUE(waiting_for_computation)) {
      stop("Current stack already computed.")
    }
    overwrite(get_output_position(), compute_value_from_current_stack())
    # NEW Modify the state
    waiting_for_computation <<- FALSE
  }
  
  list(
    current_program = function() { program },
    get_pointer_position = function() { pointer },
    get_pointer_value = get_pointer_value,
    get_stack_from_pointer = get_stack_from_pointer,
    move_pointer = function() { 
      # NEW check if the pointer can be moved
      if (!isTRUE(running)) {
        stop("Cannot move pointer: program halted.")
      }
      if (isTRUE(waiting_for_computation)) {
        stop("Cannot move pointer: operation on the current stack was not performed.")
      }
      next_position <- pointer + 4L
      if (next_position > length(program) - 1L) {
        stop("Pointer cannot be moved to position ", next_position, ".")
      }
      
      pointer <<- next_position
      # NEW change the state
      waiting_for_computation <<- TRUE
      # NEW check Opcode after moving
      check_opcode()
    },
    get_operands = get_operands,
    compute_value_from_current_stack = compute_value_from_current_stack,
    get_output_position = get_output_position,
    overwrite_output_position = overwrite_output_position,
    is_running = function() { running }
  )
}
```

I haven't yet implemented the loop. I prefer to check first if this works has expected.

Let's perform the whole tests.

Let's take the first tests. They should fail when I try to move the pointer whithout having computing the current stack:

```{r}
program <- "1,9,10,3,2,3,11,0,99,30,40,50"

process <- computer(program)

process$current_program()

test_identical <- function(...) stopifnot(identical(...))

test_identical(process$get_pointer_position(), 0L)

test_identical(process$get_pointer_value(), 1L)

test_identical(
  process$get_stack_from_pointer(), 
  c(1L, 9L, 10L, 3L)
)
```

That's it! The following test fails because I haven't computed the stack:
```{r, eval=FALSE}
process$move_pointer()
```

Now, compute the current stack.

```{r}
test_identical(process$get_output_position(), 3L)
test_identical(process$compute_value_from_current_stack(), 70L)
process$overwrite_output_position()
test_identical(
  process$current_program(),
  c(1L, 9L, 10L, 70L, 2L, 3L, 11L, 0L, 99L, 30L, 40L, 50L)
)
# Now, let's take the second operation:
process$move_pointer()
process$overwrite_output_position()
test_identical(
  process$current_program(),
  c(3500L, 9L, 10L, 70L, 2L, 3L, 11L, 0L, 99L, 30L, 40L, 50L)
)
```

Now, move to the third stack:

```{r}
process$move_pointer()
```

I should get an error, now.
```{r, eval=FALSE}
process$overwrite_output_position()
```

Yes, this works!

## Implement the loop and clean code

Now, I have to implement the loop. I don't like when programs run without a verbose mode. Otherwise I cannot debug them. I also want to be able to execute by hand the different operations. Moreover, some methods shouldn't be exposed to the user.

```{r}
computer <- function(program, auto = TRUE, verbose = TRUE) {
  waiting_for_computation <- TRUE # init
  running <- TRUE # init
  
  is_running <- function() { running }
  
  program <- eval(parse(text = paste0("c(", program, ")")))
  program <- as.integer(program)

  # moved and modified
  get_program <- function() { 
    paste(program, collapse = ",") 
  }
  
  # NEW
  if (isTRUE(verbose)) {
    message("Program succesfully read:\n", get_program())
  }
  
  pointer <- 0L # I want to respect the convention defined in the game
  

  read_position <- function(position) { 
    program[position + 1L] 
  }
  
  read_stack <- function(position) {
      stack_positions <- seq.int(position + 1L, position + 4L)
      program[stack_positions]
  }
  
  get_pointer_value <- function() { 
    read_position(pointer) 
  }

  check_opcode <- function() {
    opcode <- get_pointer_value()
    if (!(opcode %in% c(1L, 2L, 99L))) {
      stop("Wrong Opcode. Read ", opcode, " at position ", pointer, 
           ". Opcode must be 1, 2 or 99.")
    }
    
    if (identical(opcode, 99L)) {
      running <<- FALSE
      waiting_for_computation <<- FALSE
      # NEW
      if (isTRUE(verbose)) {
        message("Program halted.")
      }
    } else {
      if (isTRUE(verbose)) {
        message("Program running...")
      }
    }
  }
  
  check_opcode()

  get_operation <- function() {
    opcode <- get_pointer_value()
    switch(opcode, sum, prod)
  }
  
  get_stack_from_pointer <- function() { 
    read_stack(pointer) 
  }
  
  get_operands <- function() {
    current_stack <- get_stack_from_pointer()
    positions <- current_stack[2:3]
    read_position(positions)
  }
  
  compute_value_from_current_stack <- function() {
    operation <- get_operation()
    operands <- get_operands()
    as.integer(operation(operands))
  }
  
  overwrite <- function(position, value) {
    # NEW
    if (position + 1L > length(program)) {
      stop("Cannot write to position ", position, " because the program is of length: ", length(program), ".")
    } else {
      if (isTRUE(verbose)) {
        message("Writing value ", value, " at position ", position)
      }
    }
    new_program <- program # init
    new_program[position + 1L] <- value
    program <<- new_program # note: <<- is used. That was the motivation for a closure.
  }

  get_output_position <- function() {
    current_stack <- get_stack_from_pointer()
    current_stack[4]
  }
  
  overwrite_output_position <- function() {
    if (!isTRUE(running)) {
      stop("Cannot compute: program halted.")
    }
    if (!isTRUE(waiting_for_computation)) {
      stop("Current stack already computed.")
    }
    overwrite(get_output_position(), compute_value_from_current_stack())
    waiting_for_computation <<- FALSE
  }
  
  move_pointer <- function() { 
    if (!isTRUE(running)) {
      stop("Cannot move pointer: program halted.")
    }
    if (isTRUE(waiting_for_computation)) {
      stop("Cannot move pointer: operation on the current stack was not performed.")
    }
    next_position <- pointer + 4L
    if (next_position > length(program) - 1L) {
      stop("Pointer cannot be moved to position ", next_position, ".")
    }
    
    pointer <<- next_position
    # NEW
    if (isTRUE(verbose)) {
      message("Pointer moved at position ", pointer)
    }
    waiting_for_computation <<- TRUE
    check_opcode()
  }
  
  # NEW
  perform_next_operation <- function() {
    if (isTRUE(verbose)) {
      message("Computing...")
      message("Current stack: ", paste(get_stack_from_pointer(), collapse = ","))
    }
    overwrite_output_position()
    if (isTRUE(verbose)) {
      message("Modified program:\n", get_program())
      message("Moving pointer...")
    }
    move_pointer()
  }
  
  if (isTRUE(auto)) {
    while (is_running()) {
      perform_next_operation()
    }
  }
  
  list(
    get_program = get_program, # moved and modified
    get_pointer_position = function() { pointer },
    get_pointer_value = get_pointer_value,
    get_stack_from_pointer = get_stack_from_pointer,
    is_running = is_running, # moved
    perform_next_operation = perform_next_operation
  )
}
```

```{r}
program <- "1,9,10,3,2,3,11,0,99,30,40,50"

process <- computer(program)
```

Now, let's test with the other programs proposed by Advent of Code:

```{r}
process <- computer("1,0,0,0,99")
test_identical(process$get_program(), "2,0,0,0,99")

process <- computer("2,3,0,3,99")
test_identical(process$get_program(), "2,3,0,6,99")

process <- computer("2,4,4,5,99,0")
test_identical(process$get_program(), "2,4,4,5,99,9801")

process <- computer("1,1,1,4,99,5,6,0,99")
test_identical(process$get_program(), "30,1,1,4,2,5,6,0,99")
```

## Restoring to the "1202 program alarm" state

The final modification is to implement the "1202 program alarm" state. The trap is the indices...

```{r}
computer <- function(program, auto = TRUE, verbose = TRUE, restore_1202_state = TRUE) {
  waiting_for_computation <- TRUE # init
  running <- TRUE # init
  
  is_running <- function() { running }
  
  program <- eval(parse(text = paste0("c(", program, ")")))
  program <- as.integer(program)

  get_program <- function() { 
    paste(program, collapse = ",") 
  }
  
  # NEW
  fix_program <- function() {
    if (isTRUE(verbose)) {
      message('Restoring to the "1202 program alarm" state...')
    }
    new_program <- program # init
    new_program[2] <- 12L # position 1
    new_program[3] <- 2L # position 2
    program <<- new_program
    if (isTRUE(verbose)) {
      message("Restored program:\n", get_program())
    }
  }
  
  if (isTRUE(verbose)) {
    message("Program succesfully read:\n", get_program())
  }
  
  if (isTRUE(restore_1202_state)) {
    fix_program()
  }
  
  pointer <- 0L # I want to respect the convention defined in the game

  read_position <- function(position) { 
    program[position + 1L] 
  }
  
  read_stack <- function(position) {
      stack_positions <- seq.int(position + 1L, position + 4L)
      program[stack_positions]
  }
  
  get_pointer_value <- function() { 
    read_position(pointer) 
  }

  check_opcode <- function() {
    opcode <- get_pointer_value()
    if (!(opcode %in% c(1L, 2L, 99L))) {
      stop("Wrong Opcode. Read ", opcode, " at position ", pointer, 
           ". Opcode must be 1, 2 or 99.")
    }
    
    if (identical(opcode, 99L)) {
      running <<- FALSE
      waiting_for_computation <<- FALSE
      # NEW
      if (isTRUE(verbose)) {
        message("Program halted.")
      }
    } else {
      if (isTRUE(verbose)) {
        message("Program running...")
      }
    }
  }
  
  check_opcode()

  get_operation <- function() {
    opcode <- get_pointer_value()
    switch(opcode, sum, prod)
  }
  
  get_stack_from_pointer <- function() { 
    read_stack(pointer) 
  }
  
  get_operands <- function() {
    current_stack <- get_stack_from_pointer()
    positions <- current_stack[2:3]
    read_position(positions)
  }
  
  compute_value_from_current_stack <- function() {
    operation <- get_operation()
    operands <- get_operands()
    as.integer(operation(operands))
  }
  
  overwrite <- function(position, value) {
    if (position + 1L > length(program)) {
      stop("Cannot write to position ", position, " because the program is of length: ", length(program), ".")
    } else {
      if (isTRUE(verbose)) {
        message("Writing value ", value, " at position ", position)
      }
    }
    new_program <- program # init
    new_program[position + 1L] <- value
    program <<- new_program # note: <<- is used. That was the motivation for a closure.
  }

  get_output_position <- function() {
    current_stack <- get_stack_from_pointer()
    current_stack[4]
  }
  
  overwrite_output_position <- function() {
    if (!isTRUE(running)) {
      stop("Cannot compute: program halted.")
    }
    if (!isTRUE(waiting_for_computation)) {
      stop("Current stack already computed.")
    }
    overwrite(get_output_position(), compute_value_from_current_stack())
    waiting_for_computation <<- FALSE
  }
  
  move_pointer <- function() { 
    if (!isTRUE(running)) {
      stop("Cannot move pointer: program halted.")
    }
    if (isTRUE(waiting_for_computation)) {
      stop("Cannot move pointer: operation on the current stack was not performed.")
    }
    next_position <- pointer + 4L
    if (next_position > length(program) - 1L) {
      stop("Pointer cannot be moved to position ", next_position, ".")
    }
    
    pointer <<- next_position
    # NEW
    if (isTRUE(verbose)) {
      message("Pointer moved at position ", pointer)
    }
    waiting_for_computation <<- TRUE
    check_opcode()
  }
  
  # NEW
  perform_next_operation <- function() {
    if (isTRUE(verbose)) {
      message("Computing...")
      message("Current stack: ", paste(get_stack_from_pointer(), collapse = ","))
    }
    overwrite_output_position()
    if (isTRUE(verbose)) {
      message("Modified program:\n", get_program())
      message("Moving pointer...")
    }
    move_pointer()
  }
  
  if (isTRUE(auto)) {
    while (is_running()) {
      perform_next_operation()
    }
  }
  
  list(
    get_program = get_program,
    get_pointer_position = function() { pointer },
    get_pointer_value = get_pointer_value,
    get_stack_from_pointer = get_stack_from_pointer,
    is_running = is_running, # moved
    perform_next_operation = perform_next_operation
  )
}
```

Now, playing:

```{r}
program <- readLines("day-2/input_part-1.txt")
process <- computer(program)
```

